---
title: "two-proportion z-test"
author: "Dr Reynolds"
subtitle: "theory based approach"
---

In this section, we want to know:  for two different groups, is there a difference in the proportion for another variable.

Let's consider a case: **An experiment is run where people are basically asked "how are things going in your life this year?", but wording is slightly different: 1) "Are you having a good year?" 2) "Are you having a bad year?".  Their responses are classified as "positive" or "negative".** 

Let's answer the question:  **Is there evidence that there is an association (relationship) between the way the question is asked and how people respond**


```{r, setup = T, include=F}
library(tidyverse)
```


```{r, message=F}
read_csv(file = "https://raw.githubusercontent.com/EvaMaeRey/teaching_spring_2021/master/data/chap5.GoodandBad.csv") %>% 
  rename(wording = Wording) %>% # getting to "snake case" all lower w/ underscores
  rename(perception = Perception) ->
good_bad

good_bad
```

# Part 1. Representing bivariate (two variable), binary (two categories) associations

There are a **large** number of ways to represent associations between two binary variables.

- contingency table (or 2X2 table)
- stacked bar chart
- dodged bar chart
- mosaic plot/variable widths bar plot
- segmented bar chart (book uses a lot)

## First representation

What's the name of this first representation?

> your answer (choose 1) contingency table (or 2X2 table), stacked bar chart, dodged bar chart, mosaic plot/variable widths bar plot, segmented bar chart (book uses a lot)


```{r}
good_bad %>% 
  table() 
```

## Second representation

We can pipe the above table to plot() and get a visualization. What's it called:

> your answer (choose 1) contingency table (or 2X2 table), stacked bar chart, dodged bar chart, mosaic plot/variable widths bar plot, segmented bar chart (book uses a lot)


```{r}
good_bad %>% 
  table() %>% 
  plot()
```

## Third and fourth representations

A more common representations (in newspapers) are below...

Which do you think those are called?

> your answer (choose 2) contingency table (or 2X2 table), stacked bar chart, dodged bar chart, mosaic plot/variable widths bar plot, segmented bar chart (book uses a lot)


```{r}
ggplot(data = good_bad) +
  aes(x = wording) +
  geom_bar() + 
  aes(fill = perception)

ggplot(data = good_bad) +
  aes(x = wording) +
  geom_bar(position = "dodge") + 
  aes(fill = perception)
```

## Fifth representation used a lot in the book

We see a lot of this one in the book.  What's its name?

> your answer (choose 1) contingency table (or 2X2 table), stacked bar chart, dodged bar chart, mosaic plot/variable widths bar plot, segmented bar chart (book uses a lot)

Using the visualization, eyeball the difference in proportions.

> your answer ...

```{r}
ggplot(data = good_bad) +
  aes(x = wording) +
  geom_bar(position = "fill") + 
  aes(fill = perception)
```





## Pooled proportion

Below, we *ignore the grouping information*, looking at the overall proportion (pooled proportion) across both groups.

Estimate the pooled proportion based on the data visualization:

> your answer

This proportion is used in the calculation of the SE for the null distribution.


```{r}
ggplot(data = good_bad) +
  aes(x = "ignoring wording") +
  geom_bar(position = "fill") + 
  aes(fill = perception)
```



# Part 2. Theory based approach - How to conduct a **two-proportion z-test**


## Step 1: Statistic: difference in proportions

Our statistic is the difference in proportions for between the two groups. 

$$
\hat{p}_1 - \hat{p}_2
$$

Hide the definition above from yourself, and look only at the formula, try to describe the statistic in words:

> your answer

Look at the contingency table...

```{r}
good_bad %>% 
  table() 
```


... and use r as a calculator to calculate the value of the statistic, phat-1 (good year condition) minus phat-2 (bad year condition).  

```{r}

```


## Step 2: Characterize the distribution for the null

What are differences in proportions for the two groups that might have arisen *just by chance*? 

Standardized statistic - z score, it is hypothesized that there is usually that no difference in the proportions (pi) for these groups.  

$$
H_0: \pi_{good~framing}- \pi_{bad~framing} = 0
$$

$$
H_A: \pi_{good~framing}- \pi_{bad~framing} \ne 0
$$

Another way we describe $H_0$ this is that the variables are **statistically independent**, i.e. *knowing the value of one var doesn't tell you information about the other variable*.  

This is in contrast to **statistical dependence** -- *there is a real association between the variables*, or $H_A$ (specifically the not equal case). 

If we have enough evidence for **statistical dependence** (i.e. p-value < alpha) -- we can say the relationship is **statistically significant**!  Yea!

$$
SE = \sqrt{\hat{p}(1-\hat{p})*(\frac{1}{n_1}+\frac{1}{n_2})}
$$

Looking at the equation above use R as a calculator to calculate the standard error (SE) - the expected spread (variability) consistent with no relationship (chance model); the spread is centered at zero (no difference in proportions). 

```{r}

```



$$
z = \frac{observed~statistic-hypothesized~value}{standard~error~of~statistic}
$$

Or

$$
z = \frac{(\hat{p}_1 - \hat{p}_2)-0}{\sqrt{\hat{p}(1-\hat{p})*(\frac{1}{n_1}+\frac{1}{n_2})}}
$$

## Step 3a: Strength of evidence - zscore

Is what I observe (phat1 - phat2) consistent with the null?  Is it far enough away from the null, such that I'd say it is inconsistent? 

First we express this as a z-score. 

Show that the calculated z score is 4.30 by using R as a calculator below.  The course guide says that it will! 

```{r}
# my calculation

```



## Step 3b: Translate to z-score to pvalue 


Use `pnorm` to translate from z-score to pvalue, using only the appropriate r code (delete the 'comment character #' at the *beginning* of the line of code).  Replace the value of z with what we calculated above, or use assignment to create z above. 

```{r}
# 1 - pnorm(z) # greater than
# pnorm(z) # less than
# 2*(1-abs(pnorm(z))) # two sided
```


# Part 2. Pause. Wait a second!  Have we met the theory based validity conditions?

The validity condition for the theory based approach -- the two proportion z-test -- is that **at least 10 observations in each cell of 2X2, contingency table**.  Are the validity conditions met? Are we okay using this elegant mathematical (theory) approach, or should we kick it back to the computationally intensive simulation approach? 

```{r}
good_bad %>% 
  table() 
```


> your answer



## Part 3. Confidence interval

Note: the SE for the confidence interval is **different** than for the overall two-proportion z-test.  There is no pooling!  We see just group proportions.


$$
SE = \sqrt{\frac{\hat{p_1}(1-\hat{p_1})}{n_1}+\frac{\hat{p_2}(1-\hat{p_2})}{n_2}}
$$



$$
statistic \pm multiplier * SE~of~statistic
$$


Multiplier  (Confidence level)

* 1.645     (90%)
* 1.960     (95%)
* 2.57      (99%)


